\plaintitle

### 2017-01-08 drafts/labor/complexity.tex
complexity

\topic{Piecework's body of research most squarely addresses complexity in two of the cases we looked at earlier:
Airy's human computers and among industrial workers, especially during World War II.}
The majority of piecework's perspective, on this topic, will come from these areas.

\topic{Airy's work opened the door to greater task complexity by decomposition.}
Airy's computers had relatively limited education in mathematics, but
they were nevertheless collectively able to accomplish complex goals by
approaching distinct, subdivided, quickly verifiable tasks in chunks~\cite{grier2013computers}.
\ali{do i need to say more?}

\topic{It seemed, too, that only tasks that could be measured and priced could be completed via piecework
--- a conclusion which would fundamentally limit the potential complexity of piecework.}
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would be unsuitable for piecework~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative work like design,
which can take an arbitrary number of iterations before proceeding to a subsequent step.

\topic{The next innovation in piecework's drive toward complexity came on
the assembly line, specifically during the war effort.}
At a time when manufacturing firms needed more able--bodied workers, but
all of the conventionally trained metalworkers had joined the armed forces,
factories turned to women~\cite{honey1985creating,davies2014origins}.
This need to train workers rapidly to do very narrowly defined tasks resulted in
a \textit{recomposition} of work, wherein
workers were only prepared for the tasks they would do;
riveters would only be trained on how to do rivet and
to pass the output to the next person on the assembly line,
who could evaluate and send the work back or proceed with the next step.

\ali{OR\dots}

\topic{Only tasks that could be measured and priced could be completed via piecework
--- a fact which fundamentally limited the potential complexity of piecework.}
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would be unsuitable for piecework~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative work like design,
which can take an arbitrary number of iterations before proceeding to a subsequent step.

\topic{Piecework was limited to tasks that could be quickly and accurately evaluated.}
For example, the roles required to facilitate piecework
in the early \nth{20} century included ``piecework clerks, inspectors, and `experts'''~\cite{10.2307/23702539}.
\citeauthor{hart2016rise} argue that evaluation limited piecework's complexity:
at some point, evaluating multidimensional work for quality
(rather than for quantity) becomes infeasible.
In his words,
``if the quality of the output is more difficult to measure than the quantity [\ldots]
then a piecework system is likely to encourage
an over--emphasis on quantity~\dots~and an under--emphasis on quality''~\cite{hart2016rise}.

@@

\topic{Piecework researchers were also finding that certain types of \textit{organizations},
as well as certain types of \textit{work},
were more amenable to piecework regimes.}
\citeauthor{Brown01041990} wrote that piecework was
``less likely in jobs with a variety of duties than in
jobs with a narrow set of routinized duties''~\cite{Brown01041990};
\citeauthor{SJOE:SJOE371} hypothesized this as an effect of market forces
--- suggesting that
``pay schemes based on~\dots~performance may induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.
Airy seemed to recognize this,
decomposing complex mathematical problems into easily verified components, % ~\cite{grier2013computers},
but applying this model to tasks like railway engineering may have been prohibitively difficult.

\topic{But it was more than the inherent qualities of work and the organization of the firm that mattered;
management practices themselves had to be appropriately geared to working with pieceworkers.}
Competent foremen were integral both for allocating resources and
for liaising between workers and upper management~\cite{wray1949marginal,10.2307/2118435}.
% The West Virginia mines, as \citeauthor{10.2307/2118435} writes,
% ~\cite{10.2307/2118435}.

@@

% \clearpage
\subsubsection{\whatchanged}
\begin{comment}
  mangeerial overhead limits, so what's different
  more people can now do complex work without training (more complex)
  parts of management can be automated (more firms)
  cheaper to create the infrastructure (more complex)
\end{comment}

\topic{The research on piecework tells us that
we should expect piecework to thrive in industries where
the nature of the work is limited in complexity~\cite{Brown01041990}.}
Given the flourishing of on--demand labor platforms such as
Uber, AMT, and others, we ask ourselves
what --- if anything --- has changed.
We argue that
the internet has trivialized
the costs and challenges of the earlier limiting factors
because technology makes it easier
\begin{inlinelist}
  \item for workers to do complex work without training,
  \item to manage workers in doing complex work, and 
  \item to create the infrastructure necessary to manage the workers.
\end{inlinelist}

@@

\topic{Algorithms have automated some tasks that previously fell to management.}
Computational systems hire workers~\cite{turkitLittle,weld2010decision}, as well as direct their activities~\cite{uberAlgorithm}, and act as ``piecework clerks''~\cite{10.2307/23702539} to inspect, modify and combine work~\cite{turkopticon,takingAHITMcInnis}.
In many cases, the intermediary function has been removed as well, leading workers to need to directly email requesters for clarification and feedback~\cite{martin2014being}.
These algorithms, however, are less able than human managers to manage contingencies that were not programmed into them.

\topic{Finally, the organizational limit on infrastructure creation is somewhat lessened.}
Writing web scripts takes fewer people and fewer hours than creating physical equipment for piecework.
\citeauthor{turkitLittle}'s vision was that any user with basic programming skills could tap into on--demand human intelligence.
As better toolkits lower this threshold~\cite{myers2000past} and computational thinking diffuses, a broader population will be able to use on--demand work.

% \clearpage
\subsubsection{\implication}
Technology's ability to support human cognition will enable stronger assumptions about workers' abilities, increasing the complexity of on--demand work outcomes.
Just as the shift to expert crowdsourcing increased complexity, so too will workers with better tools increase the set of tasks possible.
Beyond this, further improvements would most likely come from replicating the success of narrowly--slicing education for expert work as \citeauthor{hart2013rise} and \citeauthor{grier2013computers} described in their piecework examples
of human computation~\cite{grier2013computers} and drastically reformulating macro--tasks given the constraints of piecework~\cite{hart2013rise}.
To some extent, an argument can be made that
MOOCs and other online education resources
provide crowd workers with the resources that they need, but 
it remains to be seen whether that work will be appropriately valued, let alone
properly interpreted by task solicitors~\cite{aguaded2013mooc}.
If we can overcome this obstacle,
we might be able to empower more of these workers to do complex work such as engineering and metalworking,
rather than doom them to ``uneducated'' match girl reputations~\cite{10.2307/3827491}.
However, many such experts are already available on platforms such as Upwork, so training may not directly increase the complexity accessible to on--demand work unless it makes common expertise more broadly available.

Will the shift from human managers to Turing--complete algorithms raise the complexity ceiling? 
By the Turing test, the algorithms would be at best indistinguishable from human piecework clerks and foremen.
So in terms of enabling coordination, algorithmic management is unlikely to directly raise the ceiling beyond what piecework could achieve.
However, as a resource constraint, algorithms are a fixed cost and not a per--person cost like human managers.
So in terms of accessibility, algorithms will allow a broader class of organizations and individuals to benefit from soliciting on--demand work.
This shift may enable complex goals that were not cost--effective before to become feasible.
However, because algorithms remain far from replicating all of the foreman's responsibilities,
most likely is a middle ground in which on--demand work re--introduces
the human element to management in a more targeted way~(e.g.,~\cite{haas2015argonaut,kulkarni2012mobileworks,crowdguilds}).
This move will require resolving the tension between workers and perilously antagonistic managers, as \citeauthor{10.2307/2118435} suggest, to break a toxic cycle of mistrustful requesters~\cite{MaliciousCrowdworkersGadiraju}.

Finally, the cost of creating piecework infrastructure has dropped. 
Expensive manufacturing equipment has been largely replaced by computer code~\cite{lessig2006code}.
As with lowered costs of management, lowered infrastructure costs will make on--demand work accessible to a broader set of people and organizations.
This in and of itself does not raise the complexity ceiling, but by broadening the potential market for on--demand work, it may enable a new set of goals and needs take part.

>>>>

\msb{ summarizing the later conclusion so we can support it here:
complexity is determined by whether we can encode it into workflows
complexity is determined by whether we can modularize and measure it
complexity is determined by how much management overhead there is
complexity is determined by the infrastructure that can support it
}

\topic{Piecework's body of research most squarely addresses complexity in two of the cases we looked at earlier:
Airy's human computers and among industrial workers.}
The majority of piecework's perspective, on this topic, will come from these areas.

\topic{Airy's work on astronomical charts opened the door to greater task complexity by encoding the intelligence into the process rather than the people.}
Airy's computers had relatively limited education in mathematics, but by combining individual simple mathematical operations, Airy was able to create a complex composite outcome~\cite{grier2013computers}.
Likewise, in Ford's factories, no individual could build the entire car, but the process could emergently produce one.

However, when piecework intially entered the American economy,
it was not used for complex work.
Without having designed processes to combine work intelligently, piecework managers were restricted to available workers' skills: it was infeasible to provide new pieceworkers with the comprehensive education
that apprenticeships imparted \cite{hart2013rise}.
So, initially piecework arose for farm work, and as
\citeauthor{hughRaynbirdTaskWork} and others discuss,
the practice remained relatively obscure until
it blossomed in the textile industry~\cite{hughRaynbirdTaskWork}.
Complexity levels remained low at the turn of
the \nth{20} century as piecework saturated New York City~\cite{riisOtherSideLives}.

\topic{Measurement also limited the complexity of piecework: only tasks that could be measured and priced could be completed via piecework.}
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers.
They identified task homogeneity and measurement as key requirements for piecework.
However, complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would then be unsuitable~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative tasks like scientific research,
which can take an arbitrary number of iterations before proceeding to a subsequent step.

%\topic{The next innovation in piecework's drive toward complexity came on
%the assembly line, specifically during the war effort.}
%At a time when manufacturing firms needed more able--bodied workers, but
%all of the conventionally trained metalworkers had joined the armed forces,
%factories turned to women~\cite{honey1985creating,davies2014origins}.
%This need to train workers rapidly to do very narrowly defined tasks resulted in
%a \textit{recomposition} of work, wherein
%workers were only prepared for the tasks they would do;
%riveters would only be trained on how to do rivet and
%to pass the output to the next person on the assembly line,
%who could evaluate and send the work back or proceed with the next step.

%\ali{OR\dots

%\topic{Only tasks that could be measured and priced could be completed via piecework
%--- a fact which fundamentally limited the potential complexity of piecework.}
%Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
%one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
%complex, creative work
%--- which is inherently heterogeneous and difficult to routinize ---
%would be unsuitable for piecework~\cite{10.2307/23702539}.
%\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
%``efficiency experts'' can effectively gauge how long known tasks should take, but
%would find themselves overwhelmed if they attempted to assess creative work like design,
%which can take an arbitrary number of iterations before proceeding to a subsequent step.

\topic{Moreover, piecework was limited to tasks that could be quickly and accurately evaluated.}
%For example, the roles required to facilitate piecework
%in the early \nth{20} century included ``piecework clerks, inspectors, and `experts'''~\cite{10.2307/23702539}.
\citeauthor{hart2016rise} argue that evaluation limited piecework's complexity:
at some point, evaluating multidimensional work for quality
(rather than for quantity) becomes infeasible.
In his words,
``if the quality of the output is more difficult to measure than the quantity [\ldots]
then a piecework system is likely to encourage
an over--emphasis on quantity~\dots~and an under--emphasis on quality''~\cite{hart2016rise}.
Complex work, which is often subjective to evaluate, falls victim to this criteria.
%``pay schemes based on~\dots~performance may induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.

@@

%\topic{Piecework researchers were also finding that certain types of \textit{organizations},
%as well as certain types of \textit{work},
%were more amenable to piecework regimes.}
%\citeauthor{Brown01041990} wrote that piecework was
%``less likely in jobs with a variety of duties than in
%jobs with a narrow set of routinized duties''~\cite{Brown01041990};
%\citeauthor{SJOE:SJOE371} hypothesized this as an effect of market forces
%--- suggesting that
%``pay schemes based on~\dots~performance may induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.
%Airy seemed to recognize this,
%decomposing complex mathematical problems into easily verified components, % ~\cite{grier2013computers},
%but applying this model to tasks like railway engineering may have been prohibitively difficult.

%\topic{But it was more than the inherent qualities of work and the organization of the firm that mattered;
%management practices themselves had to be appropriately geared to working with pieceworkers.}
%Competent foremen were integral both for allocating resources and
%for liaising between workers and upper management~\cite{wray1949marginal,10.2307/2118435}.
% The West Virginia mines, as \citeauthor{10.2307/2118435} writes,
% ~\cite{10.2307/2118435}.

@@

%}
% \clearpage
\subsubsection{\whatchanged}
\begin{comment}
  mangeerial overhead limits, so what's different
  more people can now do complex work without training (more complex)
  parts of management can be automated (more firms)
  cheaper to create the infrastructure (more complex)
\end{comment}

\topic{The research on piecework tells us that
we should expect piecework to thrive in industries where
the nature of the work is limited in complexity~\cite{Brown01041990}, and become less common as work becomes more complex.}
Given the flourishing of on--demand labor platforms such as
Uber, AMT, and others, we ask ourselves
what --- if anything --- has changed.

%We argue that
%the internet has trivialized
%the costs and challenges of the earlier limiting factors
%because technology makes it easier
%\begin{inlinelist}
%  \item for workers to do complex work without training,
%  \item to manage workers in doing complex work, and 
%  \item to create the infrastructure necessary to manage the workers.
%\end{inlinelist}

@@

\topic{Algorithms have automated some tasks that previously fell to management.}
Computational systems now %hire workers~\cite{turkitLittle,weld2010decision}, as well as direct their activities~\cite{uberAlgorithm}, and 
act as ``piecework clerks''~\cite{10.2307/23702539} to inspect, modify and combine work~\cite{turkopticon,takingAHITMcInnis}.
%In many cases, the intermediary function has been removed as well, leading workers to need to directly email requesters for clarification and feedback~\cite{martin2014being}.
These algorithms are less able than human managers to evaluate subjective work.
So, assessment remains a bottleneck.

%\topic{Finally, the organizational limit on infrastructure creation is somewhat lessened.}
%Writing web scripts takes fewer people and fewer hours than creating physical equipment for piecework.
%\citeauthor{turkitLittle}'s vision was that any user with basic programming skills could tap into on--demand human intelligence.
%As better toolkits lower this threshold~\cite{myers2000past} and computational thinking diffuses, a broader population will be able to use on--demand work.

% \clearpage
\subsubsection{\implication}
Algorithms are undoubtably capable of shepherding more complex work than the linear processes available to Airy and Ford.
However, as work becomes more complex, it becomes increasingly difficult to codify a process to achieve it~\cite{Faraj2006a,edmondson2012teaming}.
So, while algorithms will increase the complexity ceiling beyond what was possible previously with piecework, it seems likely that this increase is not enough to achieve truly complex work.

However, technology's ability to support human cognition will enable stronger assumptions about workers' abilities, increasing the complexity of on--demand work outcomes.
Just as the shift to expert crowdsourcing increased complexity, so too will workers with better tools increase the set of tasks possible.
Beyond this, further improvements would most likely come from replicating the success of narrowly--slicing education for expert work as \citeauthor{hart2013rise} and \citeauthor{grier2013computers} described in their piecework examples
of human computation~\cite{grier2013computers} and drastically reformulating macro--tasks given the constraints of piecework~\cite{hart2013rise}.
To some extent, an argument can be made that
MOOCs and other online education resources
provide crowd workers with the resources that they need, but 
it remains to be seen whether that work will be appropriately valued, let alone
properly interpreted by task solicitors~\cite{aguaded2013mooc}.
If we can overcome this obstacle,
we might be able to empower more of these workers to do complex work such as engineering,
rather than doom them to ``uneducated'' match--girl reputations~\cite{10.2307/3827491}.
However, many such experts are already available on platforms such as Upwork, so training may not directly increase the complexity accessible to on--demand work unless it makes common expertise more broadly available.

Evaluation remains as difficult for crowd work as it did for the efficiency experts. 
Reputation systems for crowdsourcing platforms remain notoriously inflated~\cite{Horton2015a}.
Creating communities of practice for peer assessment~\cite{crowdguilds} may offer a path forward, but many aspects of assessment remain subjective: whether a logo made for a client is fantastic or terrible may depend on taste.

So, in the case of complexity, the history of piecework does not yet offer compelling evidence that on--demand work will achieve far more complex outcomes. 
It will be far more broadly distributed---reaching many more tasks and areas of expertise by virtue of the internet---but not necessarily far more complex than what piecework previously achieved.

%Will the shift from human managers to Turing--complete algorithms raise the complexity ceiling? 
%By the Turing test, the algorithms would be at best indistinguishable from human piecework clerks and foremen.
%So in terms of enabling coordination, algorithmic management is unlikely to directly raise the ceiling beyond what piecework could achieve.
%However, as a resource constraint, algorithms are a fixed cost and not a per--person cost like human managers.
%So in terms of accessibility, algorithms will allow a broader class of organizations and individuals to benefit from soliciting on--demand work.
%This shift may enable complex goals that were not cost--effective before to become feasible.
%However, because algorithms remain far from replicating all of the foreman's responsibilities,
%most likely is a middle ground in which on--demand work re--introduces
%the human element to management in a more targeted way~(e.g.,~\cite{haas2015argonaut,kulkarni2012mobileworks,crowdguilds}).
%This move will require resolving the tension between workers and perilously antagonistic managers, as \citeauthor{10.2307/2118435} suggest, to break a toxic cycle of mistrustful requesters~\cite{MaliciousCrowdworkersGadiraju}.

%Finally, the cost of creating piecework infrastructure has dropped. 
%Expensive manufacturing equipment has been largely replaced by computer code~\cite{lessig2006code}.
%As with lowered costs of management, lowered infrastructure costs will make on--demand work accessible to a broader set of people and organizations.
%This in and of itself does not raise the complexity ceiling, but by broadening the potential market for on--demand work, it may enable a new set of goals and needs take part.
